---
title: "Microbial Response to Antibiotics"
author: "Kris Sankaran"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Microbial Response to Antibiotics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
link-citations: yes
bibliography: antibiotic.bib
---

## Introduction ##


```{r, setup}
library("knitr")
library("ggplot2")
library("plyr")
library("dplyr")
library("phyloseq")
library("treelapse")
opts_chunk$set(fig.width = 10, fig.height = 7, cache = FALSE)

# minimal theme for ggplots
theme_set(theme_bw())
min_theme <- theme_update(
  panel.border = element_blank(),
  panel.grid = element_blank(),
  axis.ticks = element_blank(),
  legend.title = element_text(size = 8),
  legend.text = element_text(size = 6),
  axis.text = element_text(size = 6),
  axis.title = element_text(size = 8),
  strip.background = element_blank(),
  strip.text = element_text(size = 8),
  legend.key = element_blank()
)

data(abt)
```

```{r, read-chunks}
read_chunk("~/Documents/programming/treelapse/inst/doc/antibiotic_code.R")
```

In this vignette, we study the impact of antibiotics on microbial ecology, using
data from [@dethlefsen2011incomplete]. Even non-scientists know that antibiotics
kill bacteria -- here we will try to describe the effects of these time courses
on microbial ecology in a little more precision. For example, how quickly do the
bacteria die off, does the population return to the initial states, and if so
how long does that take? Also, are certain microbes more / less resistant to
antibiotics, and do we see any kinds of competition or shifts in the the
community's position on the adaptive landscape?

We will answer these questions by studying time series of microbial abundances
while the their host goes through a few separate periods of antibiotic
treatment. The series can be arranged along the taxonomic tree; we can represent
each node by either its sum or average across descendants. Instead of
identifying each node with the full time series, we can alternatively conisder
the vector of averages during the different periods (pre, during, and
post-antibiotic) and see how these averages change across the different periods
and across different subtrees.

## Data Background ##

We give some preliminary views of the data, just to get some intuition about it.
If you are familiar with this (type of) study, you can safely skip this section.

First, we look at how many samples there are, how many microbes were found, and
what types of features were collected about the samples.

```{r, data-view}
```

Evidently, there are about fifty timepoints collected across three individuals.
Each timepoint is associated with a condition: before, between, or after
different antibiotic time courses (Cp, for ciprofloxacin). Unfortunately, the
treatments were not given at the same time for each person, see the figure
below.

```{r, sampling-times}
```

If we had time be very careful, we might want to find a way to align these
series. Instead, we will just analyze these three subjects separately.

Turning over to the RSV (microbial) abundances, we notice that the counts are
highly skewed. We're finding that the vast majority of microbes have very small
abundnaces.

```{r, skewness}
```

This is actually pretty standard in microbiome data; we'll use the filtering and
transformation functions in `phyloseq` to down to the ~300 most abundant taxa
and work with `asinh` transformed counts[^The asinh transformation is like
taking logs, but is less agressive at smaller values. This is evident from the
representation, $asinh\left(x\right) = \log\left(x + \sqrt{1 + x^{t}}\right)].

```{r, abt-transform}
```

These are relatively generic checks we would do with any microbiome data set.
Next, we consider some preparatory work specific to treelapse.

## Data Preparation ##

Before we can produce any of the `treelapse` figures, we will need to generate
the following data.

* An edgelist specifying the taxonomy of microbes in the sample.
* A `values` dataset giving the

```{r abt-values}
```

```{r get-taxa}
```

```{r aggregate-tips}
```

```{r prep-timebox-data}
```

```{r timebox-mappings}
```

## Visualization ##

### Tree Sums ###

```{r timebox}
```

```{r treebox}
```

### Tree Averages ###

```{r timebox-means-data}
```

```{r timebox-means}
```

```{r treebox-means}
```

### Per-Condition Sankey ###

```{r doi-sankey-data}
```

```{r doi-sankey}
```

## Interpretation ##

## References ##
