---
title: "Hierarchical Clustering, Parallel Coordinates, and Phyloseq"
author: "Kris Sankaran"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Hierarchical Clustering, Parallel Coordinates, and Phyloseq}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This application is motivated by
[http://statweb.stanford.edu/~susan/summer12/phyloseq-demosh.html](http://statweb.stanford.edu/~susan/summer12/phyloseq-demosh.html).

One treelapse feature illustrated here is the ability to display only those time
series associated with leaves. This is useful in this case since otherwise some
of the leaves would be overrepresentated, because many of the internal nodes
look like them and not other leaves. This option is set by the `display_opts`
parameter, setting `leaves_only = TRUE` within the `ts` parameter.

```{r setup}
library("treelapse")
library("phyloseq")
library("ape")
library("knitr")
opts_chunk$set(out.width = "700px", out.height = "600px", cache = FALSE)
data(GlobalPatterns)
gp_subset <- GlobalPatterns %>%
  filter_taxa(function(x) { var(x) > 9.22 * 1e5 }, prune = TRUE) %>%
  transform_sample_counts(asinh)
```

```{r, prep_data}
## Code from http://statweb.stanford.edu/~susan/summer12/phyloseq-demosh.html
GPUF <- UniFrac(gp_subset)
gp_hclust <- hclust(GPUF, method = "complete")

## extract edges from hierarchical clustering
edges <- as.phylo(gp_hclust)$edge
colnames(edges) <- c("parent", "child")
class(edges) <- "character"
sample_names_sub <- sample_names(gp_subset)
names(sample_names_sub) <- seq_len(nsamples(gp_subset))
rownames(edges) <- edges[, "child"]
edges[names(sample_names_sub), "child"] <- sample_names_sub

## Prepare aggregated values
X <- get_taxa(gp_subset)
taxa <- tax_table(gp_subset)
for (i in seq_len(nrow(taxa))) {
  taxa[i, ] <- na.locf(as.character(taxa[i, ]))
}
rownames(X) <- paste0("s", rownames(taxa))
plot_order <- dist(X) %>%
  hclust() %>%
  as.dendrogram() %>%
  order.dendrogram() 
X <- X[plot_order, ]

values <- tree_fun_multi(
  edges,
  X,
  tree_mean
)
values$time <- rownames(X)[values$row]

display_opts <- list(
  "x_axis_rotation" = 45,
  "axis_font_size" = 9,
  "axis_text_anchor" = "start",
  "margin" = list(
    "bottom" = 75
  ),
  "ts" = list(
    "leaves_only" = TRUE
  )
)
timebox_tree(
  values,
  edges,
  display_opts = display_opts
)
```

```{r treebox}
treebox(
  values,
  edges,
  display_opts = display_opts
)
```
