---
title: "An Introduction to the Treelapse Package"
author: "Kris Sankaran"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{DOI Trees for Microbial Abundance}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: treelapse_intro.bib
link-citations: yes
---

```{r, setup}
library("knitr")
library("ape")
library("data.table")
library("plyr")
library("dplyr")
library("treelapse")

opts_chunk$set(out.width = "450", out.height = "200px", cache = FALSE)
```

## Introduction ##

`treelapse` is an R package for interactive visualization of data that can be
arranged along trees. This interactivity is helpful simplifies navigation across
the tree (and the associated comparisons across data points).

Our motivating application is to the microbiome, where time series or
multinomial counts of microbial abundances can be arranged along
an evolutionary tree. This abstraction is useful in other domains, however.
For example, a tree can be built from geographic distances, or using
hierarchical clustering. Different applications are described in the other
vignettes to this package -- this vignette is designed just to get you started.

This package is pretty simple, since we have tried to maintain focus on a few of
the recurring questions in this problem domain. We consider three types of data,
which usually come with corresponding, relatively generic questions.

* **Scalars on trees**: Here we have a single number associated with each node in
the tree. The basic questions of interest are how the values across nodes
compare with each other, or whether there are any subtrees with consistently
large / small values. This questions are addressed by DOI trees.
* **Vectors on trees**: Sometimes each node is associated with a vector of
measurements. For example, we might have counts for treatment vs. control
groups. If we want to compare how the groups differ across subtrees, we can use
the DOI Sankeys.
* **Functions on trees**: Here, each node is associated with a function or time
series. The questions of interest are whether any interesting shapes are present
in the functions, and whether these shapes tend to be grouped along subtrees.
This question is addressed by time and treeboxes.

## Examples ##

To describe the required input format for these functions, we give a toy example
on simulated data. For more interesting motivating applications, see the other
vignettes.

In each method, we require a tree specified as an edgelist. Specifically, this
is a two column `data.frame`, with two columns, `parent` and `child`, specifying
the edge structure in the tree. We simulate such an edgelist using the `ape`
package. The tip nodes are numbered 1 to 100, and all others are internal nodes.
The root is node 101.

```{r, make-edgelist}
n_tips <- 100
tree <- rtree(n_tips)
edges <- tree$edge
class(edges) <- "character"
colnames(edges) <- c("parent", "child")
```

Now, we need to associate data with each node. In general, this will be a
`data.frame`, with the two columns,

* `unit`: The name of the node to which the value is associated. The name must
be the same as the name in `edges`.
* `value`: The scalar value associated with the node.

The DOI Sankeys require an extra column, `group`, for each node, while the
time and treebox require an extra column `time`. Hence, each value is associated
with either a `(unit)`, `(unit, group)` or `(unit, time)` tuple (i.e., we expect
"tall" and not "wide" representations). This is illustrated below.

### DOI Trees ###

DOI trees expect just a scalar associated with each node. Here, we generate a
random uniform variable for each node.

```{r, doi-values}
units <- unique(as.character(edges))
values <- data.frame(
  "unit" = units,
  "value" = runif(n_tips + tree$Nnode)
)
```

The result is displayed below. Clicking on a node increases its
"degree-of-interest" (DOI), and refocuses the display on that node. Further,
previously hidden nodes are revealed depending on their proximity to this new
focus node.

The parameters `leaf_width` and `size_min` control the spacing and size of
points -- `leaf_width` controls the $x$ spacing between nodes, while `size_min`
specifies the radius of the smallest node.

```{r, doi-display-1}
doi_tree(
  values,
  edges,
  style_opts = list("leaf_width" = 50, "size_min" = 2)
)
```

Suppose we didn't have data for each internal node. It might be interesting to
aggregate up the tree. This is what the `tree_sum` function does. There are
analogous functions for computing averages and aggregate diversities.

```{r, doi-tip-values}
# reduce to just tips
tip_values <- values %>%
  filter(unit %in% 1:n_tips)
tip_names <- tip_values$unit
tip_values <- setNames(tip_values$value, tip_names)

# aggregate tip values
summed_values <- tree_sum(edges, tip_values)
values <- data.frame(
  unit = names(summed_values),
  value = summed_values
)
```

The resulting tree is somewhat more pleasing to look at, because width of an
edge leading into a node is equal to the sum of the widths flowing out (since
we aggregated values from the tips up).

```{r, doi-display-2}
doi_tree(
  values,
  edges,
  style_opts = list(
    "leaf_width" = 50,
    "size_min" = 2
  )
)
```

### DOI Sankeys ###

For the DOI Sankey, we include a grouping variable. We will suppose there are
tip values for each of three groups, and then aggregate up the tree one group at
a time. The function below generates random tip values and does this
aggregation. Notice that we have to convert the unit and group columns to
characters: by default these will be integers, and `doi_sankey` will fail.

```{r, make-sankey-data}
make_grouped_data <- function(n_groups, n_tips, edges) {
  grouped_values <- matrix(
    runif(n_tips * n_groups), n_tips, n_groups,
    dimnames = list(seq_len(n_tips), seq_len(n_groups))
  )

  grouped_list <- list()
  for (i in seq_len(n_groups)) {
    grouped_list[[i]] <- tree_sum(edges, grouped_values[, i])
  }

  grouped_values <- do.call(rbind, grouped_list) %>%
    melt(varnames = c("group", "unit"))
  grouped_values$unit <- as.character(grouped_values$unit)
  grouped_values$group <- as.character(grouped_values$group)
  grouped_values
}
```

Given this data, we can create the sankey visualization. Clicking on edges
will turn focus to the "node" directly above the click-on edge. Of course, here
the groups all have about the same width, because the leaves have randomly
generated values.

```{r, vis-sankey-doi}
grouped_values <- make_grouped_data(3, n_tips, edges)
doi_sankey(grouped_values, edges)
```

### Time / Treeboxes ###

The input format for time and treeboxes is similar to that for the DOI sankey,
where we have multiple measurements of the same unit across multiple timepoints
(rather than across groups). In fact, we can use the same function to generate
toy data.

```{r, make-timebox-data}
n_times <- 100
timebox_values <- make_grouped_data(100, n_tips, edges)
colnames(timebox_values) <- c("time", "unit", "value")
```

Below we display the timebox. Here, pressing "new box" will let you draw a brush
across the time series, highlighting any that pass through the box. The
corresponding tree nodes will also be highlighted, and their names are displayed
on mouseover. Further pressing "new box" will let you draw another brush over
the time series -- only time series passing through both boxes will be
highlighted. The display in the top right is a miniature view of the main time
series display that can be used for zooming in on certain time / value windows.
Click and drag a brush over this sub-display to refocus the main time series
display the associated time and value ranges. Finally, searching for a node name
in the search box will highlight the associated series and node in red.

```{r vis-timebox}
style_opts <- list("size_min" = 1)
timebox_tree(timebox_values, edges, style_opts = style_opts)
```

Timeboxes only let you query based on time series shapes. Treeboxes are the
natural converse -- they let you select nodes in the tree by drawing brushes
over them. Other than this, the interaction is exactly the same.

```{r vis-treebox}
treebox(timebox_values, edges, style_opts = style_opts)
```

## Previous Work ##

Few of the ideas in this package are new. The key conceptual advances behind DOI
trees, DOI sankeys, and time / treeboxes were introduced in
[@heer2004doitrees], [@hochheiser2002visual], and [@guerra2013visualizing],
respectively. Further, the implementation here would be impossible without the
htmlwidgets package [@vaidyanathan2014htmlwidgets] -- it's only through this
this package that we can link interactive data analysis in R with interactive
visualization using D3.

## References ##
